Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*

### 2020-09-04 paper.Rmd
small edit in conclusion
September 1-4 smoothing

Blair and Sambanis [-@blair:sambanis:2020, hereafter B\&S] argue that theory is essential for creating models that have high accuracy in forecasting civil war onset. Indeed they assert that with such theory forecasting is more accurate than has previously been possible. Setting aside the validity of this argument, we re-examine the empirical basis for the claims made in their article. We find that these claims are false. Their theory-based escalation model does not do better than the alternatives that they examine. Actually, it does worse. The reason for this reversal of their conclusion is that they have made several mistakes in their research procedure. The performance results they report are based on smoothed performance curves, not the original unsmoothed curves. Further, two of the structural alternatives to their basic escalation model were incorrectly implemented. We also found that the scoring of their forecasts for the first half of 2016 were incorrectly performed using civil war incidence, not onset. In what follows, we show the impact of these mistakes on the conclusions.

B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars "remarkably accurately". Indeed, B&S argue that this so-called theoretical model outperforms four other "more mechanical" alternatives. Second, they claim that the integration of structure with process is better over short forecasting windows. Third, they preregistered the list of thirty countries that have the highest risk of civil war onset. They claim that such prospective predictions are rare in the literature when, in fact, they have been routine for many years with several prominent projects. B\&S claim to be unique in assessing these forecasts. A qualitative analysis of their predictions allows them to conclude that their model is robust. We will return to their analysis later, after correcting the procedural mistakes we found in their research process.

@@

B&S' analysis is based on the use of non-linear, non-deterministic machine learning models, and specifically random forests, one of which has a specification they argues is theory-based, and several others with more generic sets of covariates. Notably, the analysis is at the country-month or country-(6 month) levels and relies in large part on indicators derived from the ICEWS event data. For the moment, we set aside a) the logic of this hypothesis and b) whether their model has more theory than is typically found in empirical conflict models. In short, they uphold their assumption that theory-guided empirical research produces better conflict predictions than machine learning inspired efforts that are necessarily ad hoc combinations of available variables. They arrive at this conclusion by examining the problem of predicting civil war onset. They report that a parsimonious model using a small number of covariates derived from escalation theories of conflict can forecast civil war onset better than alternative specifications based on generic covariates not specifically informed by theory, including a \textit{kitchen sink} model with more than 1,000 covariates.

@@

- Escalation: a set of ten indicators, putatively drawn from a theoretical escalation model, for interactions between the government on one side and opposition or rebel actors on the other. 
- Quad: ICEWS quad counts, i.e. material conflict, material cooperation, verbal conflict, verbal cooperation, for interactions between the government and opposition or rebels. These are directed, thus making for four directed dyads, which with four quad categories make sixteen covariates. 
- Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from the ICEWS data for the same four directed dyads, for a total of four covariates. 
- CAMEO: counts for all CAMEO event codes over the four actor dyads, totaling $1,160$ covariates, which are mostly zero for any country in any month. 
- Average: unweighted average of the predictions from the four models briefly described above.

@@

Finally, B&S used their escalation model to create forecasts for the first half of 2016, and in their third and final analysis, they score the forecasts accuracy using civil war onset data later observed. This is summarized in B&S Table 3.

@@

The most consequential issue that we found is that all AUC-ROC values reported in B&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, empirical ROC curves. The data in rare events problems like this one restrict the number of true positive rate values and lead to non-continuous ROC curves; B&S refer to smoothing in the context of their Figure 1 with ROC curves as easing interpretation. But in fact smoothing is used to produce all AUC-ROC values they report. Standard practice in conflict research and forecasting has been to use empirical, not smoothed ROC curves, both for visualization and when calculating AUCs.

@@

B&S intend that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set. However, they actually weight the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set.^[See `1mo_run_escalation_weighted_PITF.R` line 4, where the PITF predictions are taken from the training data set (`train$pred_prob_plus1`). The next line is a hack extending the shorter `weight` vector with missing values to avoid a R warning when it is multiplied with the longer vector of escalation model test set predictions. Similarly, in the 6-month version of this file.] This appears to be an easily corrected coding error on the part of B&S.

@@

This description suggests that B&S intended to run two separate random forest models, one each on the low- and high-risk training data splits. The replication code does indeed run two separate random forecasts, but they both utilize the *exact same training data*, which consists of the full training data from all other models. In short, rather than using only data for high-risk countries to train their split-sample model for high-risk cases, the data they use captures both high- and low-risk countries, similarly for the training set for their low-risk split.

@@

Further, the AUC-ROC values reported in the original B&S Table 2 are calculated based on slightly different numbers of underlying test set cases (see Table \ref{tab:table2-N}). ROC calculations for a set of predictions can only be done on the set of cases for which both non-missing predictions and non-missing outcomes are available. Those sets differ across models (columns) for each row in B&S Table 2.

Thus a difference in AUC-ROC values for two models could be because they were calculated on different sets of underlying cases, not because the models are systemically performing at a different level. In other words, the results for different models in B&S Table 2 are not comparable to one another, and any conclusions drawn from such comparison are potentially incorrect.

@@

We now turn to an examination of our analysis that addresses and fixes the issues discussed above.^[The code for all of our analysis undertaken for this effort may be found at https://github.com/andybega/Blair-Sambanis-replication.] The main results of the original analysis consist of the comparison of the Escalation to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B&S Table 1), and a comparison of the Escalation model to models that add structural variables/information (our Table \ref{tab:table2-fixed}, B&S Table 2). We will review the substantive implications of our updated analysis below, but the bottom line is that these changes turn B&S's conclusions on their heads.

@@

This implies that the ROC curves were only smoothed in the referenced Figure 1. However, this is not the case. All AUC-ROC calculations throughout the replication code use an option to smooth the ROC curves prior to AUC calculation. ROC curves are constructed from the false positive and true positive rates as one moves through a set of ranked predictions, and as a result they appear step-like. Figure \ref{fig:rocs} shows our replication of both the estimated smoothed ROC curves from the B&S report (left-hand side) and the actual empirical ROC curves (right-hand side). The standard method is to compute the area under the curve (AUC) statistic on the original, empirical ROC curves that are shown on the right.

@@

The specific predictions also include groups of cases with identical predicted probabilities, which accounts for the unusual diagonal lines seen in the panels on the right. In any case, with a sparse outcome like civil war onsets, the true positive rate on the *y*-axis only changes when the prediction for an observed positive case is reached. For these ROC curves, and for that matter in the fundamental train/test split used for 12 of the 18 rows/models in B&S Table 1, there are only 11 civil war onset cases in the test set. Thus, the ROC curves here are very step-like, with only 12 (11 positive cases plus 1 for TPR = 0) distinct *y* coordinates. Notice also that the smoothing averages the left-most almost straight line with the right-most almost straight line in a monotonic way.

@@

What impact did the ROC smoothing have overall on the performance of the Escalation model relative to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B&S Table 1) and structural extensions (our Table \ref{tab:table2-fixed}, B&S Table 2)?  Figure \ref{fig:benefit-plot} shows the changes in AUC-ROC values had we used smoothed ROC curves to calculate the AUC-ROC values. Each point corresponds to the change in AUC-ROC values for one of the models in the cells in Tables \ref{tab:table1-nosmooth} and \ref{tab:table2-fixed} (the colors match those in Figure \ref{fig:rocs}). The vertical bar in each plot marks the average effect of smoothing on AUC-ROC. For all alternative models, from Quad to "PITF Only", smoothing sometimes hurts and sometimes benefits, but the overall impact is negligible in average. The Escalation model _always_ benefits from smoothing, with an average improvement on the order of 0.05. This is sufficient to push the Escalation model ahead of the alternative models, and accounts for the result B&S find, namely that the escalation model is generally superior.

@@

Table \ref{tab:table1-nosmooth} (B&S Table 1) shows the comparison of the Escalation model to other alternative models based on ICEWS event data indicators. The AUC-ROC values are based on original, non-smoothed ROC curves. 
Both the Average and CAMEO models outperform the Escalation model in almost all instances (see also Figure \ref{fig:table1-plot}). The Goldstein model generally outperforms the Escalation model in the 6-month version. The Quad model appears to be roughly on par with the Escalation model. Thus, the original B&S conclusion that the Escalation model is superior to the alternative models is entirely conditional on the non-standard use of smoothed ROC curves and overturns when using conventional AUC-ROC calculations.

@@

The PITF split-population analog still performs well, but the simple Escalation + PITF predictors model arguable performs better still.

@@

Adding PITF predictors actually improves performance in both cases; the "With PITF Predictions" model strictly dominates the "Escalation Only" model, and by quite a margin in the 6-month data version.

@@

The weighted model performs roughly on par with the Escalation Only model. One finding that remains is that the "PITF Only" model is outperformed by the "Escalation Only" model. As the former only uses annual inputs, but the data at hand are the 1-month or 6-months level, this is neither surprising, nor noteworthy.

@@

B&S report (Table 4)confusion matrices for their forecasts for the first half of 2016 (2016-H1). To create the confusion matrices, B&S treat the 30 highest ranked predictions as positive predictions ("1s") and the rest as negative predictions ("0s"). We replicate their Table 4 in the top portion of \ref{tab:table4}. There are two confusion matrices for slightly different codings of outcomes in 2016-H1, under the corresponding "Assuming Persistence" and "Assuming Change" headings. We should note that the "Assuming Persistence" corresponds to the values in the replication data; the small variations for the "Assuming Change" version are hand-coded in the replication code file that generates the confusion matrices and appear to be subjective assessments of B&S.

We can see that the original table presents 15 or 16 positive cases for 2016-H1, depending on the dependent variable coding variation. This corresponds to a positive rate of around 9.5% for the first half of 2016 data. In contrast, the corresponding 6-month version of the data from 2001 to 2015, with 30 half-years, has in total 20 civil war onset events, for a much lower positive rate of around 0.5%. The positive event rate in the confusion matrices far exceeds the rate of observed civil war onsets in both the training and test data. This suggests that the forecasts were erroneously assessed using civil war incidence, not onset. By examining the replication code we were able to verify that the forecasts were scored using civil war incidence, i.e., including ongoing civil wars, rather than civil war onset years only.

@@

The correct confusion matrices when using observed onset (or the lack of it) are shown in the second part of Table \ref{tab:table4}. In the default "Assuming Persistence" coding, there are no civil war onsets in the data for 2016-H1. Thus, the recall values is undefined, while the precision is 0/30 = 0, compared to reported recall and precision values of 13/15 = 0.87 and 13/30 = 0.43. The alternative coding ("Assuming Change") produces two civil war onsets. Recall is 1.0 compared to 14/16 = 0.88 before, and precision is 2/30 = 0.07 instead of 14/30 = 0.47.

Another, minor issue or rather coding error, is related to using a lead version of the DV. With the lead version of the DV, "incidence_civil_ns_plus1", which is what the models are predicting, the predicted value for 2016-H1 indicates the risk of civil war onset in 2016-H2. In the Table 4 script referenced above, the 2016-H1 predictions (for 2016-H2) are assessed using the raw DV, "incidence_civil_ns", not the lead version. Essentially, the forecasts for 2016-H2 are assessed using observed outcomes for 2016-H1. In this case, it doesn't make a difference since both the raw DV and lead version for 2016-H1 do not have any positive values.

@@

We encountered several issues in the code underlying the B&S analysis. The problems we encountered are not subjective modeling choices. When we fix these issues and perform an updated analysis, the B&S conclusions are all essentially overturned. In other words, B&S findings are based on a faulty analysis, and invalid.

@@

- The theory-driven escalation model is outperformed both by the low-effort 1,160 predictor all-CAMEO model.
- The Average ensemble model and the CAMEO models outperform the escalation model in all instances.
- Adding structural variables substantially improves the escalation model's performance.

>>>>

Blair and Sambanis [-@blair:sambanis:2020, hereafter B\&S] argue that theory is essential for creating models that have high accuracy in forecasting civil war onset. Indeed they assert that with such theory forecasting is more accurate than has previously been possible. Setting aside the validity of this argument, we re-examine the empirical basis for the claims made in support of it. We find that these claims to be unsupported and evidence presented for them to be incorrect. Their theory-based escalation model does not do better than the alternatives that they examine. Actually, it does worse. The reason for this is that they have made several mistakes in their research procedure. Further, the performance results they report are based on smoothed performance curves, not the original unsmoothed curves. This provides misleading results. In addition, two of the structural alternatives to their basic escalation model were incorrectly implemented. We also found that the scoring of their forecasts for the first half of 2016 were incorrectly performed using civil war incidence, not onset. In what follows, we show the impact of these mistakes on the conclusions.

B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars "remarkably accurately". Indeed, B&S argue that this theoretical model outperforms four other "more mechanical" alternatives.\footnote{We set aside for here the characteristics of theoretical models, accepting the B\&S denotation.}  They also claim that the integration of structure with process is better than alternaties over short forecasting windows. Third, they preregistered the list of thirty countries that have the highest risk of civil war onset. They claim that such prospective predictions are rare in the literature when, in fact, they have been routine for many years with several prominent projects. B\&S claim to be unique in assessing these forecasts. A qualitative analysis of their predictions allows them to conclude that their model is robust. We will return to their analysis later, after correcting the procedural mistakes we found in their research process.

@@

B&S' analysis is based on the use of non-linear, non-deterministic machine learning models, and specifically random forests, one of which has a specification they argue is theory-based.  Several others with more generic sets of covariates are also developed and compared. Notably, the analysis is at the country-month or country-(6 month) levels and relies in large part on indicators derived from the ICEWS event data. In short, they uphold their assumption that theory-guided empirical research produces better conflict predictions than machine learning inspired efforts that are necessarily ad hoc combinations of available variables. They arrive at this conclusion by examining the problem of predicting civil war onset. They report that a parsimonious model using a small number of covariates derived from escalation theories of conflict can forecast civil war onset better than alternative specifications based on generic covariates not specifically informed by theory, including a \textit{kitchen sink} model with more than 1,000 covariates.

@@

- Escalation: a set of ten indicators, drawn from a theoretical escalation model, for interactions between the government on one side and opposition or rebel actors on the other. 
- Quad: ICEWS quad counts, i.e. material conflict, material cooperation, verbal conflict, verbal cooperation, for interactions between the government and opposition or rebels. These are directed, thus making for four directed dyads, which with four quad categories make sixteen covariates. 
- Goldstein: -10 (conflictual) to 10 (cooperative) scores derived from the ICEWS data for the same four directed dyads, for a total of four covariates. 
- CAMEO: counts for all CAMEO event codes over the four actor dyads, totaling $1,160$ covariates, which are mostly zero for any country in any month. 
- Average: unweighted average of the predictions from the four models briefly described above.

@@

Finally, B&S used their escalation model to create forecasts for the first half of 2016. In their third analysis, they score the forecasts accuracy using subsequent civil war onset data. This is summarized in B&S Table 3.

@@

The most consequential issue that we found is that all AUC-ROC values reported in B&S Tables 1 and 2 are calculated using smoothed ROC curves, not the original, empirical ROC curves. The data in rare events problems like this one restrict the number of true positive rate values and lead to non-continuous ROC curves; B&S refer to smoothing in the context of their Figure 1 with ROC curves justified as easing interpretation. Smoothing is used to produce all AUC-ROC values they report. There is no difficulty to be overcome in using non-smoothed plots. Indeed, such plots are standard practice in conflict research and forecasting, both for visualization and when calculating AUCs.

@@

We believe that B&S intend that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set. However, they actually weight the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set.^[See `1mo_run_escalation_weighted_PITF.R` line 4, where the PITF predictions are taken from the training data set (`train$pred_prob_plus1`). The next line is a hack extending the shorter `weight` vector with missing values to avoid a R warning when it is multiplied with the longer vector of escalation model test set predictions. Similarly, in the 6-month version of this file.] This appears to be an easily corrected coding error on the part of B&S.

@@

This description suggests that B&S intended to run two separate random forest models, one each on the low- and high-risk training data splits. The replication code does indeed run two separate random forecasts, but they both utilize the *exact same training data*, which consist of the full training data from all other models. In short, rather than using only data for high-risk countries to train their split-sample model for high-risk cases, the data they use captures both high- and low-risk countries, similarly for the training set for their low-risk split.

@@

Further, the AUC-ROC values reported in the original B&S Table 2 are calculated  on slightly different numbers of underlying test set cases (see Table \ref{tab:table2-N}). ROC calculations for a set of predictions can only be done on the set of cases for which both non-missing predictions and non-missing outcomes are available. Those sets differ across models (columns) for each row in B&S Table 2.

Thus a difference in AUC-ROC values for two models could be because they were calculated on different sets of underlying cases, not because the models are systemically performing at a different level. In other words, the results for different models in B&S Table 2 are not comparable to one another, and conclusions drawn from such comparison are potentially incorrect.

@@

We now turn to an examination that addresses and fixes the issues discussed above.^[The code for all of our analysis undertaken for this effort may be found at https://github.com/andybega/Blair-Sambanis-replication.] The main results of the original analysis consist of the comparison of the Escalation to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B&S Table 1), and a comparison of the Escalation model to models that add structural variables/information (our Table \ref{tab:table2-fixed}, B&S Table 2). We will review the substantive implications of our updated analysis below, but the bottom line is that these changes turn B&S's conclusions on their heads.

@@

This suggest that using smoothed ROC curves will have no substantive impact on the results; however, in this smoothing determines the comparative results.
Their statement also implies that the ROC curves were only smoothed in the referenced Figure 1. However, this is not the case. All AUC-ROC calculations throughout the replication code use an option to smooth the ROC curves *prior* to AUC calculation. ROC curves are constructed from the false positive and true positive rates as one moves through a set of ranked predictions, and as a result they appear step-like. Figure \ref{fig:rocs} shows our replication of both the estimated smoothed ROC curves from the B&S report (left-hand side) and the actual empirical ROC curves (right-hand side). The standard method is to compute the area under the curve (AUC) statistic on the original, empirical ROC curves that are shown on the right.

@@

The specific predictions also include groups of cases with identical predicted probabilities, which accounts for the unusual diagonal lines seen in the panels on the right. In any case, with a sparse outcome like civil war onsets, the true positive rate on the *y*-axis only changes when the prediction for an observed positive case is reached. For these ROC curves, and for that matter in the fundamental train/test split used for 12 of the 18 rows/models in B&S Table 1, there are only 11 civil war onset cases in the test set. Thus, the ROC curves here are very step-like, with only 12 (11 positive cases plus 1 for TPR = 0) distinct *y* coordinates. Notice also that the smoothing averages the left-most almost straight line with the right-most almost straight line in a monotonic way.  Thus, smoothing changes the ordering of alternatives. It is not a cosmetic change as suggested.

@@

What impact did the ROC smoothing have overall on the evaluated performance of the Escalation model relative to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B&S Table 1) and structural extensions (our Table \ref{tab:table2-fixed}, B&S Table 2)?  Figure \ref{fig:benefit-plot} shows the changes in AUC-ROC values had we used smoothed ROC curves to calculate the AUC-ROC values. Each point corresponds to the change in AUC-ROC values for one of the models in the cells in Tables \ref{tab:table1-nosmooth} and \ref{tab:table2-fixed} (the colors match those in Figure \ref{fig:rocs}). The vertical bar in each plot marks the average effect of smoothing on AUC-ROC. For all alternative models, from Quad to "PITF Only", smoothing sometimes hurts and sometimes benefits, but the overall impact is negligible in average. The Escalation model _always_ benefits from smoothing, with an average improvement on the order of 0.05. This is sufficient to push the Escalation model ahead of the alternative models, and accounts for the incorrect result B&S report, namely that the escalation model is generally superior.

@@

Table \ref{tab:table1-nosmooth} (B&S Table 1) shows the comparison of the Escalation model to other alternative models based on ICEWS event data indicators. The AUC-ROC values are based on original, non-smoothed ROC curves. 
Both the Average and CAMEO models outperform the Escalation model in almost all instances (see also Figure \ref{fig:table1-plot}). The Goldstein model generally outperforms the Escalation model in the 6-month version. The Quad model appears to be roughly on par with the Escalation model. Thus, the original B&S conclusion that the Escalation model is superior to the alternative models is entirely conditional on the non-standard use of smoothed ROC curves.

@@

The PITF split-population analog still performs well, but the simple Escalation + PITF predictors model arguably performs better still.

@@

Adding PITF predictors actually improves performance in both cases; the "With PITF Predictions" model strictly dominates the "Escalation Only" model by quite a margin in the 6-month data version.

@@

The weighted model performs roughly on par with the Escalation Only model. One finding that remains is that the "PITF Only" model is outperformed by the "Escalation Only" model. As the former only uses annual inputs, but the data at hand are the 1-month or 6-months level, this is neither surprising, nor especially noteworthy.

@@

B&S report (in their Table 4) confusion matrices for their forecasts for the first half of 2016 (2016-H1). To create the confusion matrices, B&S treat the 30 highest ranked predictions as positive predictions ("1s") and the rest as negative predictions ("0s"). We replicate their Table 4 in the top portion of \ref{tab:table4}. There are two confusion matrices for slightly different codings of outcomes in 2016-H1, under the corresponding "Assuming Persistence" and "Assuming Change" headings. We should note that the "Assuming Persistence" corresponds to the values in the replication data; the small variations for the "Assuming Change" version are hand-coded in the replication code file that generates the confusion matrices and appear to be subjective assessments of B&S.

We can see that the original table presents 15 or 16 positive cases for 2016-H1, depending on the dependent variable coding variation. This corresponds to a positive rate of around 9.5% for the first half of 2016 data. In contrast, the corresponding 6-month version of the data from 2001 to 2015, with 30 half-years, has in total 20 civil war onset events, for a much lower positive rate of around 0.5%, i.e., an order of magnitude smaller. The positive event rate in the confusion matrices far exceeds the rate of observed civil war onsets in both the training and test data. This suggests that the forecasts were erroneously assessed using civil war incidence, not onset. By examining the replication code we were able to verify that the forecasts were scored using civil war incidence, i.e., including ongoing civil wars, rather than civil war onset years only.

@@

The correct confusion matrices when using observed onset (or the lack of it) are shown in the second part of Table \ref{tab:table4}. In the default "Assuming Persistence" coding, there are no civil war onsets in the data for 2016-H1. Thus, the recall value is undefined, while the precision is 0/30 = 0, compared to reported recall and precision values of 13/15 = 0.87 and 13/30 = 0.43. The alternative coding ("Assuming Change") produces two civil war onsets. Recall is 1.0 compared to 14/16 = 0.88 before, and precision is 2/30 = 0.07 instead of 14/30 = 0.47.\footnote{Another, minor issue or rather coding error, is related to using a lead version of the DV. With the lead version of the DV, "incidence_civil_ns_plus1", which is what the models are predicting, the predicted value for 2016-H1 indicates the risk of civil war onset in 2016-H2. In the Table 4 script referenced above, the 2016-H1 predictions (for 2016-H2) are assessed using the raw DV, "incidence_civil_ns", not the lead version. Essentially, the forecasts for 2016-H2 are assessed using observed outcomes for 2016-H1. In this case, it doesn't make a difference since both the raw DV and lead version for 2016-H1 do not have any positive values.}

@@

We encountered several issues in the code underlying the B&S analysis. The problems we encountered are not subjective modeling choices. When we fix these issues and perform an updated analysis, the B&S conclusions are all essentially overturned. In other words, B&S findings are based on a faulty analysis, invalid, and should be disregarded.

@@

- The theory-driven escalation model is outperformed both by the low-effort 1,160 predictor all-CAMEO model and the ensemble average model.
- The Average ensemble model and the CAMEO models outperform the escalation model in all instances.
- Adding structural variables substantially improves the escalation model's performance.
